Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-12 paper/sections/abstract.tex
added text.
[ paper ] perliminaries added.
Paper (#36)
* minor changes.

* [ paper ] section 2 improvements.

* [ paper ] major changes, abstract, conclusions was added.

* [ paper ] minor changes, title and abstract.

* [ paper ] fixing some mistakes.

* [ paper ] fixing error grammars and adding notes.

* [ paper ] section 3 introduction added.

* [ paper ] appendix added.

* added text.

* [ paper ] perliminaries added.

* [ paper ] implementation added, some proofs added.

* [ paper ] fixing some typos and text added.

* [ paper ] fixing some typos and text added.

* [ paper ] complexity m

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda, a  proof assistant for type theory, solutions obtained by
\Metis, an automatic theorem prover, to problems in classical 
propositional logic. The translation method consists of justifying 
step-by-step each deduction found in the prover's solution. To carry 
out this plan, we emulate the inference rules of the prover, 
providing a version of them in the proof assistant. Finally, we 
parse the output of the prover, build and analyze the respective 
tree associated with the proof, for ending up with an \Agda code of 
the proof for type-checking.

\keywords{proof reconstruction
\unsure{proof reconstruction or proof-reconstruction},
\Agda, automatic theorem prover, classical propositional logic, \Metis.}

>>>>

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
the proof assistant for type theory \Agda, solutions obtained by
the  automatic theorem prover \Metis to problems in classical
propositional logic. The translation method consists of justifying
step-by-step each deduction found in the prover's solution. To carry
out this plan, we emulate the inference rules of the prover,
providing a version of them in the proof assistant. Finally, we
parse the output of the prover, build and analyze the respective
tree associated with the proof, for ending up with an \Agda code of
the proof for type-checking.

\keywords{proof-reconstruction, \Agda, automatic theorem prover,
classical propositional logic, \Metis.}

### 2017-10-12 paper/sections/introduction.tex
[ paper ] diagram added.
[ paper ] fixed athena process description.

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related
work, Section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows.
We briefly mention notation and syntax of the classical propositional
logic in Section \ref{sec:preliminaries}.
In Section \ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in Section
\ref{sec:conclusions}. At the end, we list some related works in
proof reconstruction in Section \ref{sec:related-work}.
The appendix \textsc{A} shows a complete example of the proof
reconstruction workflow.

>>>>

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013,Bohme2011,Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related
work, Section \ref{sec:related-work}.

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality. The formalization of these rules allowed us to justify step-by-step \Metis' \TSTP derivations by translating automatically to \Agda by a proof-reconstruction tool built for this research named \Athena.
We structure the paper as follows.
We briefly mention some notation and syntax of the classical propositional logic in Section~\ref{ssec:CPL}.
In Section~\ref{ssec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in
Section~\ref{sec:conclusions}. Some related works with this paper is presented in Section~\ref{sec:related-work}.
Finally, we provide a complete example in the appendix \textsc{A} of the proof reconstruction workflow.

### 2017-10-12 paper/sections/metis-language-and-proofs.tex
[ paper ] fixed athena process description.

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality, that allows us to justify step-by-step the proofs delivered in
\TSTP format generated by the prover \Metis.

@@

@@

Finally, \emph{axiom} rule appears in \TSTP derivation by one of the following rules to in the problem a \verb!fof! axiom: \canonicalize, \clausify, \conjunct, or \simplify. Then, the last rule remains is \emph{resolve}, it appears by the
binary rule, \resolve, the version in \Metis of the resolution
theorem. For a more detailed description of this rules see the Section~\ref{sec:proof-reconsruction}.

\subsubsection{Proofs.}
\label{ssec:metis-proofs}

>>>>

@@

\begin{figure}
\centering
  \begin{bprooftree}\tt
    \AxiomC{}
    \RightLabel{negate}
    \UnaryInfC{$\neg p$}
    \RightLabel{strip}
    \UnaryInfC{$\neg p$}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$p$}
    \RightLabel{canonicalize}
    \UnaryInfC{$p$}
    \RightLabel{simplify}
    \BinaryInfC{$⊥$}
    \RightLabel{canonicalize}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \caption{The \Metis' proof tree for $p \vdash p$ from the
  derivation in Fig.~\ref{fig:metis-proof-tstp}}
  \label{fig:metis-example}
\end{figure}

@@

Finally, \emph{axiom} rule appears in \TSTP derivation by one of the following rules to in the problem a \verb!fof! axiom: \canonicalize, \clausify, \conjunct, or \simplify. Then, the last rule remains is \emph{resolve}, it appears by the
binary rule, \resolve, the version in \Metis of the resolution
theorem. For a more detailed description of this rules see the Section~\ref{sec:proof-reconstruction}.

\subsubsection{Proofs.}
\label{sssec:metis-proofs}

### 2017-10-12 paper/sections/preliminaries.tex
Paper (#36)
* minor changes.

* [ paper ] section 2 improvements.

* [ paper ] major changes, abstract, conclusions was added.

* [ paper ] minor changes, title and abstract.

* [ paper ] fixing some mistakes.

* [ paper ] fixing error grammars and adding notes.

* [ paper ] section 3 introduction added.

* [ paper ] appendix added.

* added text.

* [ paper ] perliminaries added.

* [ paper ] implementation added, some proofs added.

* [ paper ] fixing some typos and text added.

* [ paper ] fixing some typos and text added.

* [ paper ] complexity measure added.

* [ paper ] minor changes.

Before describe the proof reconstruction, we need some definitions
and some convention notations.

@@

\textit{Propositions.}
We define the classical propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$),
binary logic connectives ($∧, ∨, ⇒, ⇔)$,
propositional negation $(¬)$, and logic constants $⊥$
and $⊤$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.
%Then, $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
$⊥, ⊤ \in \Prop$.
If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where $(φ\ \square\ ψ)$ stands for a proposition that use the $\square$ symbol as one of the binary connectives mentioned above.
If $φ \in \Prop$, then $(¬ φ) \in \Prop$.

@@

\begin{equation*}
\label{eq:derivation}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro}
  \BinaryInfC{$φ ∧ ψ$.}
\end{bprooftree}
% \qquad
% \begin{bprooftree}
%   \AxiomC{$[¬ φ]$}
%   \noLine
%   \UnaryInfC{$\vdots$}
%   \noLine
%   \UnaryInfC{$⊥$}
%   \RightLabel{\abbre{RAA}}
%   \UnaryInfC{φ}
% \end{bprooftree}
\end{equation*}

@@

\begin{equation*}
\label{eq:PEM-RAA}
\begin{bprooftree}
\AxiomC{}
\RightLabel{PEM}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$,}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{$Γ, ¬ φ ⊢ ⊥$}
\RightLabel{RAA}
\UnaryInfC{$Γ ⊢ φ$.}
\end{bprooftree}
\end{equation*}

@@

\emph{Proof Trees.} We use trees to represent derivations and
natural deduction proofs (see for instance, the proof tree in a
\Metis' derivation in Fig.~\ref{fig:metis-example}).
In a proof tree, the root is the conclusion of the entire
derivation. The nodes labeled with the name of the inference rule
contains the result of applying that rule to their parents. Lastly,
the leaves in the tree become the premises, not all necessary
canceled (see a better discussion about canceling  hypotheses in proof trees in Section 2.4 in \citeauthor{VanDalen1994}~\cite{VanDalen1994}).

>>>>

@@

\textit{Propositions.}
We define the classical propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$),
binary logic connectives ($∧, ∨, ⇒, ⇔)$,
propositional negation $(¬)$, and logic constants $⊥$
and $⊤$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.
\begin{itemize}
\item $⊥, ⊤ \in \Prop$.
\item $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
\item If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where the $\square$ symbol is one of the binary connectives mentioned above.
\item If $φ \in \Prop$, then $(¬ φ) \in \Prop$.
\end{itemize}

@@

\begin{equation*}
\label{eq:derivation}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro.}
  \BinaryInfC{$φ ∧ ψ$}
\end{bprooftree}
% \qquad
% \begin{bprooftree}
%   \AxiomC{$[¬ φ]$}
%   \noLine
%   \UnaryInfC{$\vdots$}
%   \noLine
%   \UnaryInfC{$⊥$}
%   \RightLabel{\abbre{RAA}}
%   \UnaryInfC{φ}
% \end{bprooftree}
\end{equation*}

@@

\begin{equation*}
\label{eq:PEM-RAA}
\begin{bprooftree}
\AxiomC{}
\RightLabel{PEM,}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{$Γ, ¬ φ ⊢ ⊥$}
\RightLabel{RAA.}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
\end{equation*}

@@

\emph{Proof Trees.} We use trees to represent derivations and
natural deduction proofs (see for instance, the proof tree in a
\Metis' derivation in Fig.~\ref{fig:metis-example}).
In a proof tree, the root is the conclusion of the entire
derivation. The nodes labeled with the name of the inference rule
contains the result of applying that rule to their parents. Lastly,
the leaves in the tree become the premises, not all necessary
canceled.
 % (see a better discussion about canceling  hypotheses in proof trees in Section 2.4 in \citeauthor{VanDalen1994}~\cite{VanDalen1994}).

### 2017-10-12 paper/sections/proof-reconstruction.tex
[ paper ] fixed athena process description.

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types is valid (see a description of \Agda for its main features in
Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in Step~3 find
a solution if the problem is a theorem, in that case \Metis replies
with a derivation in \TSTP format of the proof, otherwise the problem is counter feasible.

% With such a derivation, in Step~5, we process the derivation with our \Haskell translator tool, \Athena~\cite{Athena}.
% \Athena parses the \TSTP format, analyze the
% derivation and generates a representation of the natural deduction
% proof using a tree data structure (see the properties
% of this tree in Section~\ref{ssec:metis-proofs}).
% We perform an analysis of the proof-tree removing redundancies,
% unnecessary steps or unused output from the \TSTP derivation.

% As a result,
% \Athena generates an \Agda file with the proof-term. We have included
% in this file, \emph{imports} that call our \Agda libraries that give
% support for CPL presented in~\cite{AgdaProp}, and support for \Metis'
% inference rules implemented in~\cite{AgdaMetis}.

% Finally, we use the proof checker in Step~7 to type-check the
% proof-term. If type-checking succeed, the \TSTP
% derivation is correct, otherwise when it fails, one of the three
% actors in the process could be the responsible: \Metis, \Athena, or
% \Agda.

>>>>

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types is valid (see a description of \Agda for its main features in
Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can reply with a derivation in \TSTP format of the proof.

With the \TSTP  derivation from the Step~2, we process the derivation with the \Athena translator tool~\cite{Athena}.
\Athena parses the \TSTP format, analyzes the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Section~\ref{sssec:metis-proofs}).
We perform an analysis of the proof-tree to remove some
unnecessary steps that introduce redundancies or unused input for
\Athena.

As a result of using \Athena, we get an \Agda file, the proof-term. To generate this file, \Athena has used names of functions and theorems from \Agda libraries:
\texttt{agda-prop}~\cite{AgdaProp}, and \texttt{agda-metis}~\cite{AgdaMetis}.

Finally, we type-check the \Agda proof-term. If type-checking succeed, the \TSTP derivation delivered by \Metis was correct.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be investigated by the user, checking \Metis, \Athena, or even \Agda.
